---
title: "Developing Data Products - Assignment #1"
author: "Lena Horsley"
date: "2/18/2020"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Assignment
Using R Markdown, create a web page featuring a map created with Leaflet and hosted on GitHub Pages, RPubs, or NeoCities.

### Data Set Creation
For this assignment, I decided to create a leaflet map of Pokemon Go landmarks (gyms, nests, habitats, and trainer spots) for several cities in North and South Carolina. These resources were used to create the data set:

* [Pokemon Go Trading Wiki (Pokemon Go Wiki) - North Carolina](https://pkmngotrading.com/wiki/North_Carolina)
* [Pokemon Go Trading Wiki (Pokemon Go Wiki) - South Carolina](https://pkmngotrading.com/wiki/South_Carolina)
* [Top Places to Catch Pokémon in Myrtle Beach](https://www.myrtlebeach.com/things-to-do/top-places-to-catch-pokmon-in-myrtle-beach/)
* [Pokemon Go Gyms in Asheville](http://ashevilleblog.com/pokemon-go-gyms-in-asheville/)
* [GPS Coordinates](https://www.gps-coordinates.net/)

Each landmark is denoted by a circle marker, where the color is representative of the landmark type. After clicking on the marker, a popup appears containing:

* location name
* address
* city and county
* GPS coordinates

```{r plot1, echo = TRUE}
suppressPackageStartupMessages(library(leaflet))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(here))

options(digits = 3)
set.seed(7779311)
theme_set(theme_minimal())
pokemon <- here("/testData/NC_SC_PokemonGoData.csv") %>%
  read_csv()
head(pokemon)
pal <- colorFactor(
  palette = c('red3', 'purple', 'springgreen4','black'),
  domain = pokemon$type
)

leaflet(data = pokemon) %>%
  addTiles() %>%
  addCircleMarkers(clusterOptions = markerClusterOptions(), 
                   color = ~pal(type), 
                   popup = paste(pokemon$location, "<br>",
                           pokemon$address, "<br>",
                           pokemon$city, pokemon$county, "<br>",
                           pokemon$latitude, pokemon$longitude)) %>%
  addLegend(pal = pal, values = ~type)
```


